\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-11-29 Manuscript/manuscript.tex
PUB

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach, which will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the first case, one can have a full explanation of the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable.
Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the process theory is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. The present work is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset.

@@

In this section we are going to review the theoretical concepts on which the probabilistic methods here used are based. We will expose them following the approaches of Murphy \cite{murphy2012machine},  Koller et al. \cite{koller2009probabilistic}, H{\o}jsgaardand \cite{hojsgaard2012graphical} et al.  and Russel et al. \cite{russell2010artificial}.  Furthermore we will provide also a rapid overview about the main concepts of information theory,  following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches,  since we used some of its concepts in the preliminary analysis of the dataset. Finally this chapter will be concluded with a rapid overview about the concept of algorithm intepretability. Such concept, together with the accuracy and the confusion matrix, is necessary for the assessment of a ML performances. On the other side the concepts related to the celestial mechanics here used will be described, following the Murray approach \cite{murray1999solar} into the Appendix A

\section{Markov random fields}

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
where V is the number of variables and 1:V stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms the transition probability tensors \cite{wu2017markov}.  As one can point out, the number of parameters is cumbersome as the number of variables grows: indeed the number of parameter required scales as $\mathcal{O}(K^{V})$.  Such formidable problem can be attacked by considering the concept of conditional independence. This is defined as\cite{murphy2012machine}:

@@

In this case a first order Markov chain is obtained,  where the transition tensor is of second order \cite{wu2017markov}.  Given this formalism we are interested in finding a smart way to plot such joint distribution into an intuitive way: the graph theory provide the answer to this quest.  In particular the random variables can be represented by nodes and presence of conditional independence for two random variables by the lack of an edge that interconnects them.  Bayesian networks consider directed edges,  while Markov random fields (MRF) only undirected.  As consequence,  while the the concept of topological ordering,  by which the parents n nodes are labelled with a lower  with respect to their children,  is well defined for Bayesian network,  for MRF is not.  In order to solve this issue it is useful to consider the Hammersley-Clifford theorem as stated in \cite{murphy2012machine}:

@@

Finally in order to reduce the computational cost,  one can consider only the pairwise interaction instead of the maximum clique. This is the analogue of what is usually performed in solid state physics (but surely not always)  when only the interaction between the first neighbour is considered.  Another example is the  Ising model: here we have a lattice of spins that can be or in $\ket{+}$ or in $\ket{-}$ and their interaction is modelled by\cite{murphy2012machine}:

@@

where $w_{st}=J$ represent the coupling strength between two neighbour site.  The collective state is described by

@@

where $\otimes$ is the tensor product.  If this parameter is associated with a positive finite value we have an associative Markov network: basically collective states in which all sites have the same configuration is favoured. Thus we will have two collective states: one for which we have all $\ket{+}$ and another in which we have all $\ket{-}$  Such situation would model,  in principle,  the ferromagnet materials where the external magnetic field induce into the material a magnetic filed with the same direction.  On the other side if the magnetization of the material is opposite with respect to the external field,  and thus $J<0$,  we have an anti-ferromagnetic system in which frustrated states are present.  Furthermore lets consider the unnormalized log probability of a collective state $\textbf{y}=\ket{i_{1},i_{2},...,i_{n}}$ \cite{murphy2012machine}:

@@

But this is nothing more that the well know \footnote{In physics} Hamiltonian of an Ising system. This is not a simple coincidence:  indeed the Hamiltonian of a system represent,  rudely speaking,  its total energy.  Thus according to the Boltzmann or Gibbs distribution we have \cite{murphy2012machine}:

@@

From this expression it is clear why this method is called moment matching; it is worth nothing tha such computation is largely expensive from a computational point of view: thus scholar usually consider other techinques or at least stochastic gradient descent method. A full review can be found in \cite{murphy2012machine} and \cite{koller2009probabilistic}. Finally we consider, as for the dataset in analysed in this work, the case where we have both discrete and continuous variables i.e. $x=\left(i_{1},...,i_{d},y_{1},...,y_{q} \right)$ with d discrete variable and q continuos variables. The are called in the literature Mixed Interaction Models. In this case the following density has to be considered \cite{hojsgaard2012graphical}:

@@

Moreover one can further modify the previous form in order to obtain a particular factorial expansion: such models are referred as homogeneous mixed interaction models \cite{hojsgaard2012graphical}.

\section{Information theory}
Given an ensemble of random variable, we can qunaitify the amount of information that one variable contains of another one: such quantity is called mutual information and it is a key concept within the information theory. This approach, that was implemented by Claude Shannon decades before the probabilistic modelling, represent a complementary way by which one can attack the problem of conditional dependence between random variables. Here we would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow to properly define the concept of mutual information. The founding concept of information theory is the entropy. This quantity express the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficently the second one describe the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated to another cannot have a degree of uncertainty lower with respect to the quantity that its aimed to describe. On these basis we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
As for the joint distribution also in this case we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
Finally we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chains, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise we would have a clear violation of the second principle of thermodynamics (see for instance the Maxwell's demon \cite{feynman2018feynman})

@@

Beside the accuracy and other features that are connected with the confusion matrix, another important feature is the algorithm intepretability. This feature, at a first instance, is connected on how much the algorithm provide explanation to a human on how it choose a forecast instead of another one, or in other word how much the algorithm is not a blackbox. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g the right to explanation)}, in the author opinion results provided by such algorithm cannot be considered scientific, since the scientific method also requires to provide an explanation and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the a formal theory T can be translated into S if and only if S is able to proof the theorem of T in its own language. On the other side we would also that an algorithm, as an explanation, is complete: we expect that it will be also able to make correct forecasts for all available data. However as shown by \cite{doshi2017towards} and by \cite{gilpin2018explaining} the intepretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal system as algorithms or scientific theories. For instance lets consider the fall of object: at first glance one can consider the only the items on Earth, in this case the acceleration is constant and it is a very simple theory; but as one moves to consider also the interaction within planets a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete, the second one more difficult to explain but much more complete. Therefore, in principle, the evaluation of an algorithm performances should be considered not only in to one dimension, the accuracy, but also by considering its intepretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into an more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research center. Among the 40 features that were present in the dataset we excluded the ones that were clearly redundant (such as the distance quantity evaluated in miles instead of kilometres), the ones that were connected only to the other name of the asteroid, the one connected to the name of the orbit and the one connected with the orbiting planet, since for all it was the Earth. Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A, since a description of them cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only the 16 \% of the asteroids were hazardous, we considered to reduce the number of the non-hazardous in order to improve the portion of the hazardous ones. For this purpose we constructed a database in which the proportion hazardous/not hazardous was 1:5: thus all the hazardous one were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle the proportion should be much less unbalanced, however the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here used was a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

\begin{table}[]
\caption{The features used for the present analysis. The unit of measure are reported in the original dataset \cite{kaggle_dataset}.}
\begin{center}
\begin{tabular}{c|c}
\hline
\textbf{Features}             & \textbf{Type}        \\ \hline
Neo Reference ID              & not used             \\ \hline
Absolute Magnitude            & Continuous           \\ \hline
Est Dia in KM (min)           & Continuous           \\ \hline
Est Dia in KM (max)           & Continuous           \\ \hline
Close Approach Date           & Continuous           \\ \hline
Epoch Date Close Approach     & Continuous           \\ \hline
Relative\_Velocity            & Continuous           \\ \hline
Miss\_Dist                    & Continuous           \\ \hline
Min\_Orbit\_Intersection      & Continuous           \\ \hline
Jupiter\_Tisserand\_Invariant & Continuous           \\ \hline
Epoch\_Osculation             & Continuous           \\ \hline
Eccentricity                  & Continuous           \\ \hline
Semi Major Axis               & Continuous           \\ \hline
Inclination                   & Continuous           \\ \hline
Asc Node Longitude            & Continuous           \\ \hline
Orbital Period                & Continuous           \\ \hline
Perihelion Distance           & Continuous           \\ \hline
Perihelion Arg                & Continuous           \\ \hline
Perihelion Time               & Continuous           \\ \hline
Mean\_Anomaly                 & Continuous           \\ \hline
Mean\_Motion                  & Continuous           \\ \hline
Hazardous                     & Categorical (Binary)
\end{tabular}
\end{center}
\label{tab_features}
\end{table}

@@

This chapter is organized in the following way: first we are going to report the results concerning the preliminary analysis performed on the dataset. This include the factor analysis of mixed data and the mutual information analysis of the continuous variables of asteroids vs their hazard. Then it will follow the analysis of the dataset performed with the probabilistic methods: after a preliminary analysis on the continuous variables, the mixed interaction model as well the minforest model obtained for the whole dataset will be presented and discussed. Finally the probabilistic models previously obtained will be compared with the outputs and the performances of four machine learning algorithm (Random Forest, Support vector machines Quadratic Discriminant Analysis and Logistic regression).

@@

We started the analysis with the graphical models by inspecting the relations between continuous variables in the dataset: thus in this first step the binary variable \textit{Hazardous} was excluded. For this purpose, among the different methods available, we considered the \textit{Graphical least absolute shrinkage and selection operator} GLASSO as implemented in the \textit{glasso} R package \cite{friedman2008sparse,glasso}. After different tests reported in Fig. \ref{GLASSO_convergence}, we considered as final result the graph obtained with the value of $\rho$ (the one that penalize further connections) equal to $0.3$. This is because, in this plot, we see that the conditional dependences/independences stated by the Celestial Mechanics are correctly reproduced: for instance we see that the diameter of the asteroids is independent from all features related to its motion. This is definitely meaningful since the mass of the asteroids is largely lower with respect to the mass of earth: thus there is no way by which the asteroid orbit can be modified by its mass as explained in Appendix A. Furthermore we see that the Close approach Date and the Epoch date are dependant each other but in no way from the other features: this is right since the date and epoch are set with an arbitrary scale. Also the Perihelion Epoch and Osculation Time are dependant, as expected, but there is no dependence with the orbit parameters. We see that the mean motion is conditionally independent with respect to relative velocity, but correctly this is dependent from perihelion distance.
\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.1.pdf}
    \subcaption{$\rho$=0.1}
    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.2.pdf}
    \subcaption{$\rho$=0.2}
    \vspace{4ex}
  \end{minipage} \\
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.3.pdf}
    \subcaption{$\rho$=0.3}
    \vspace{4ex}
  \end{minipage}%%
    \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.4.pdf}
    \subcaption{$\rho$=0.4}
    \vspace{4ex}
  \end{minipage}%%
\caption{GLASSO analysis, performed with the glasso package \cite{friedman2008sparse,glasso}, with different $\rho$ parameter (the one that penalize further connections) for the Asteroid dataset without the discrete variable Hazardous. The plots were obtained with the \textit{igraph} package for R \cite{igraph}}
\label{GLASSO_convergence}
\end{figure}

>>>>

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning.
The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. The present work is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset.

@@

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
Where V is the number of variables and 1:V stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms, the transition probability tensors \cite{wu2017markov}.  As one can point out, the number of parameters is cumbersome as the number of variables grows: indeed, the number of parameter required scales as $\mathcal{O}(K^{V})$.  Such a challenging problem can be attacked by considering the concept of conditional independence. This is defined as\cite{murphy2012machine}:

@@

In this case, a first-order Markov chain is obtained,  where the transition tensor is of second-order \cite{wu2017markov}.  Given this formalism, we are interested in finding an intelligent way to plot such joint distribution intuitively: the graph theory answers this quest.  In particular, the absence of an edge that interconnects the random variables can be represented by nodes and the presence of conditional independence for two random variables.  Bayesian networks consider directed edges,  while Markov random fields (MRF) are only undirected.  Consequently,  while the concept of a topological ordering,  by which the parents n nodes are labelled lower to their children,  is well defined for the Bayesian network, MRF is not.  In order to solve this issue, it is helpful to consider the Hammersley-Clifford theorem as stated in \cite{murphy2012machine}:

@@

Finally, to reduce the computational cost,  one can consider only the pairwise interaction instead of the maximum clique. This is the analogue of what is usually performed in solid-state physics (but not always)  when only the interaction between the first neighbour is considered.  Another example is the  Ising model: here we have a lattice of spins that can be or in $\ket{+}$ or in $\ket{-}$ and their interaction is modelled by\cite{murphy2012machine}:

@@

Where $w_{st}=J$ represent the coupling strength between two neighbour site.  The collective state is described by

@@

Where $\otimes$ is the tensor product, if this parameter is associated with a positive finite value, we have an associative Markov network: collectively, all sites with the same configuration are favoured. Thus we will have two collective states: one for which we have all $\ket{+}$ and another in which we have all $\ket{-}$  Such a situation would model,  in principle,  the ferromagnetic materials where the external magnetic field induces into the material a magnetic field with the same direction.  On the other side, if the material's magnetisation is opposite to the external field,  and thus $J<0$,  we have an anti-ferromagnetic system in which frustrated states are present.  Furthermore lets consider the unnormalized log probability of a collective state $\textbf{y}=\ket{i_{1},i_{2},...,i_{n}}$ \cite{murphy2012machine}:

@@

The previous expression is the Hamiltonian of an Ising system. This is not a simple coincidence:  indeed, the Hamiltonian of a system represent,  rudely speaking,  its total energy.  Thus according to the Boltzmann or Gibbs distribution, we have \cite{murphy2012machine}:

@@

From this expression, it is clear why this method is called moment matching; it is worth noting that such computation is largely expensive from a computational point of view: thus, scholars usually consider other techniques or at least the stochastic gradient descent method. A full review can be found in \cite{murphy2012machine} and \cite{koller2009probabilistic}. Finally we consider, as for the dataset in analysed in this work, the case where we have both discrete and continuous variables i.e. $x=\left(i_{1},...,i_{d},y_{1},...,y_{q} \right)$ with d discrete variable and q continuous variables. They are called in the literature Mixed Interaction Models. In this case, the following density has to be considered \cite{hojsgaard2012graphical}:

@@

Moreover, one can further modify the previous form in order to obtain a particular factorial expansion: such models are referred to as homogeneous mixed interaction models \cite{hojsgaard2012graphical}.

\section{Information theory}
Given an ensemble of random variables, we can quantify the amount of information that one variable contains: such quantity is called mutual information and is a key concept within information theory. This approach, which was implemented by Claude Shannon decades before the probabilistic modelling, represents a complementary way by which one can attack the problem of conditional dependence between random variables. Here we would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow defining the concept of mutual information properly. The founding concept of information theory is entropy. This quantity expresses the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficiently the second one describes the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated with another cannot have a degree of uncertainty lower to the quantity it aims to describe. On these bases, we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
As for the joint distribution also in this case, we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
Finally, we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chain, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise, we would have a clear violation of the second principle of thermodynamics (see for instance, the Maxwell's demon \cite{feynman2018feynman}).

@@

Besides the accuracy and other features connected with the confusion matrix, another critical feature is algorithm interpretability. This feature represents how much the algorithm explains its predictions and, in general, its mechanics. In other words, how much the algorithm is not a black box. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g. right to explanation)}, in the author opinion, results provided by such an algorithm cannot be considered scientific since the scientific method also requires to explain and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the formal theory T can be translated into S if and only if S can prove the theorem of T in its language. On the other side, we would also know that an algorithm, as an explanation, is complete: we expect that it will also be able to make correct forecasts for all available data. However, as shown by \cite{doshi2017towards}, and by \cite{gilpin2018explaining} the interpretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal systems as algorithms or scientific theories. For instance, let us consider the fall of object: at first glance, one can consider the only the items on Earth, in this case, the acceleration is constant, and it is a straightforward theory; but as one moves to consider also the interaction within planets a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete; the second one is more difficult to explain but much more complete. Therefore, in principle, the evaluation of algorithm performances should be considered not only into one dimension, the accuracy but also by considering its interpretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since their description cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, we considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, we constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

\begin{table}[]
\caption{The features used for the present analysis. The units of measure are reported in the original dataset \cite{kaggle_dataset}. The explanation of these features is provided in Appendix A. }
\begin{center}
\begin{tabular}{c|c}
\hline
\textbf{Features}             & \textbf{Type}        \\ \hline
Neo Reference ID              & not used             \\ \hline
Absolute Magnitude            & Continuous           \\ \hline
Est Dia in KM (min)           & Continuous           \\ \hline
Est Dia in KM (max)           & Continuous           \\ \hline
Close Approach Date           & Continuous           \\ \hline
Epoch Date Close Approach     & Continuous           \\ \hline
Relative\_Velocity            & Continuous           \\ \hline
Miss\_Dist                    & Continuous           \\ \hline
Min\_Orbit\_Intersection      & Continuous           \\ \hline
Jupiter\_Tisserand\_Invariant & Continuous           \\ \hline
Epoch\_Osculation             & Continuous           \\ \hline
Eccentricity                  & Continuous           \\ \hline
Semi Major Axis               & Continuous           \\ \hline
Inclination                   & Continuous           \\ \hline
Asc Node Longitude            & Continuous           \\ \hline
Orbital Period                & Continuous           \\ \hline
Perihelion Distance           & Continuous           \\ \hline
Perihelion Arg                & Continuous           \\ \hline
Perihelion Time               & Continuous           \\ \hline
Mean\_Anomaly                 & Continuous           \\ \hline
Mean\_Motion                  & Continuous           \\ \hline
Hazardous                     & Categorical (Binary)
\end{tabular}
\end{center}
\label{tab_features}
\end{table}

@@

This chapter is organized in the following way: first, we will report the results concerning the preliminary analysis performed on the dataset. The factor analysis of mixed data and the mutual information analysis of the continuous variables of asteroids vs their hazard will be reported. Then it will follow the analysis of the dataset performed with the probabilistic methods. After a preliminary analysis on the continuous variables, the mixed interaction model and the minforest model obtained for the whole dataset will be presented and discussed. Finally, the probabilistic models previously obtained will be compared with the outputs and the performances of four machine learning algorithms (Random Forest, Support vector machines, Quadratic Discriminant Analysis and Logistic regression).

@@

The analysis with the graphical models was started by inspecting the relations between continuous variables in the dataset. Therefore, in this first step, the binary variable \textit{Hazardous} was excluded. For this purpose, among the different methods available, we considered the \textit{Graphical least absolute shrinkage and selection operator} GLASSO as implemented in the \textit{glasso} R package \cite{friedman2008sparse,glasso}. After different tests reported in Fig. \ref{GLASSO_convergence}, we considered as the final result the graph obtained with the value of $\rho$ (the one that penalizes further connections) equal to $0.3$. Such choice is motivated by the fact that the conditional dependences/independences stated by the Celestial Mechanics are correctly reproduced with this parameter. For instance, we see that the diameter of the asteroids is independent of all features related to their motion. Such a result is meaningful since the volume/mass of the asteroids is largely lower than the mass of the earth.  Thus, as explained in Appendix A, there is no way the asteroid orbit can be modified by its mass. In addition, we see that the Close approach Date and the Epoch date are linked to each other but in no way from the other features. This result is correct since the date and epoch are set with an arbitrary scale. Also, as expected, the Perihelion Epoch and Osculation Time are linked, but there is no dependence on the orbit parameters. Furthermore, the mean motion is conditionally independent with relative velocity, but correctly, this is dependent on perihelion distance.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.1.pdf}
    \subcaption{$\rho$=0.1}
    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.2.pdf}
    \subcaption{$\rho$=0.2}
    \vspace{4ex}
  \end{minipage} \\
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.3.pdf}
    \subcaption{$\rho$=0.3}
    \vspace{4ex}
  \end{minipage}%%
    \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/GLASSO_0.4.pdf}
    \subcaption{$\rho$=0.4}
    \vspace{4ex}
  \end{minipage}%%
\caption{GLASSO analysis, performed with the glasso package \cite{friedman2008sparse,glasso}, with different $\rho$ parameter (the one that penalize further connections) for the Asteroid dataset without the discrete variable Hazardous. The plots were obtained with the \textit{igraph} package for R \cite{igraph}}
\label{GLASSO_convergence}
\end{figure}
